---
title: "Stat 435 Lecture Notes 5 & 5a: Practice"

header-includes:
   - \usepackage{bbm}
   - \usepackage{amssymb}
   - \usepackage{amsmath}
   - \usepackage{graphicx}
   - \usepackage{natbib}
   - \usepackage{float}
   - \floatplacement{figure}{H}
output:
  pdf_document: default
fontsize: 11pt
---

```{r, echo=FALSE, warning=FALSE}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

# Diagnosis for Logistic regression

```{r, cache=T, echo=T, warning=F, fig.height=6,fig.width=6.5}

library(ISLR)
Hitters= na.omit(Hitters)
Hitters$Salary1 = as.numeric(Hitters$Salary > 500)

fitRes = glm(Salary1~Years,data = Hitters, family = "binomial")
par(mfrow = c(2,2))
plot(fitRes)

# check residuals; caution: these residuals are not as desired
max(fitRes$residuals)
min(fitRes$residuals)


# plot residuals versus fitted values
par(mfrow = c(1,1)) # row-column
OrdinaryRes = Hitters$Salary1 - fitRes$fitted.values 
plot(fitRes$fitted.values,OrdinaryRes,
     main="Ordinary residuals vs estimated probabilites")
# add a lowess fit
lines(lowess(OrdinaryRes, fitRes$fitted.values,f=0.8), col=2)

# plot logit of estimated probabilities against predictors
plot(Hitters$Years,fitRes$fitted.values,
     main="Predictor vs estimated probabilites")
abline(lm(fitRes$fitted.values ~ Hitters$Years))

```

## Extract more statistics from `glm`

First, let us check what `groom::augment` applied to a `glm` object does by reading https://rdrr.io/cran/broom/man/augment.glm.html.

When newdata is not supplied, augment.lm returns one row for each observation, with seven columns added to the original data: 

```{r, eval=F}

.hat 
Diagonal of the hat matrix
 
.sigma 
Estimate of residual standard deviation when corresponding observation is dropped from model
 
.cooksd 
Cooks distance, cooks.distance()
 
.fitted 
Fitted values of model
 
.se.fit 
Standard errors of fitted values
 
.resid 
Residuals
 
.std.resid 
Standardised residuals
```

Some unusual lm objects, such as rlm from MASS, may omit .cooksd and .std.resid. gam from mgcv omits .sigma. 

When newdata is supplied, returns one row for each observation, with three columns added to the new data: 

```{r,eval=F}
.fitted 
Fitted values of model
 
.se.fit 
Standard errors of fitted values
 
.resid 
Residuals of fitted values on the new data
 


```
  
## Illustration on extracting statistics for diagnosis

The following are adopted from: www.sthda.com/english/articles/36-classification-methods-essentials/148-logistic-regression-assumptions-and-diagnostics-in-r/

```{r, cache=T, echo=T, warning=F,,fig.height=6,fig.width=6.5}
library(dplyr)
library(tidyverse)
library(broom)
theme_set(theme_classic())

# Load the data
data("PimaIndiansDiabetes2", package = "mlbench")
PimaIndiansDiabetes2 <- na.omit(PimaIndiansDiabetes2)
# Fit the logistic regression model
model <- glm(diabetes ~., data = PimaIndiansDiabetes2,family = binomial)
# Predict the probability (p) of diabete positivity
probabilities <- predict(model, type = "response")
predicted.classes <- ifelse(probabilities > 0.5, "pos", "neg")
head(predicted.classes)

# Select only numeric predictors
mydata <- PimaIndiansDiabetes2 %>% dplyr::select_if(is.numeric) 
predictors <- colnames(mydata)
# Bind the logit and tidying the data for plot
mydata <- mydata %>%
  mutate(logit = log(probabilities/(1-probabilities))) %>%
  gather(key = "predictors", value = "predictor.value", -logit)

# example for "gather": https://www.rdocumentation.org/packages/tidyr/versions/0.8.3/topics/gather

ggplot(mydata, aes(logit, predictor.value))+
  geom_point(size = 0.5, alpha = 0.5) + geom_smooth(method = "loess") + 
  theme_bw() +  facet_wrap(~predictors, scales = "free_y")


# Extract model results
model.data <- augment(model) %>% mutate(index = 1:n()) 

# augment: add columns to the original data that was modeled
# info on what is augmented: https://rdrr.io/cran/broom/man/augment.glm.html

model.data %>% top_n(3, .cooksd)

ggplot(model.data, aes(index, .std.resid)) + 
  geom_point(aes(color = diabetes), alpha = .5) +
  theme_bw()

## plot cook's distance vs estimated probabilities
plot(model.data$index,model.data$.cooksd, main="cook's distance vs observation number")


```

# Diagnostics for multinomial logistic regression

```{r, cache=T, echo=T, warning=F}
library(glmnet)
par(mfrow = c(1,1)) 
load("MultinomialExample.RData")

# find the best lambda for lasso
cvfit=cv.glmnet(x, y, family="multinomial", alpha=1, parallel = TRUE)
plot(cvfit)
# extract fitted model
cvmd = cvfit$glmnet.fit

# extract estimated ceofficients
tmp_coeffs <- coef(cvfit, s = "lambda.min")
tmp_coeffs[[1]]@x
 
# obtain predictions
#Type '"class"' applies only to
#  '"binomial"' or '"multinomial"' models, and produces the
#  class label corresponding to the maximum probability.
  
prdProb = predict(cvfit, newx = x, s = "lambda.min", type = "response")
prdClass = predict(cvfit, newx = x, s = "lambda.min", type = "class")

```

```{r, cache=T, echo=T, warning=F}
library(glmnet)
library(ggplot2)
head(iris)
tempcv <- cv.glmnet(x=as.matrix(iris[,-5]), y=iris[,5], family="multinomial", 
                    nfolds=20, alpha=1)
coefsMin <- coef(tempcv, s="lambda.min")
# show coefficients
coefsMin

```

# Poisson regression


```{r, cache=T, echo=T, warning=F,}
require(ggplot2)

p <- read.csv("poisson_sim.csv")
  p$prog <- factor(p$prog, levels=1:3, labels=c("General", "Academic", 
                                                     "Vocational"))
summary(p)


m1 <- glm(num_awards ~ prog + math, family="poisson", data=p)



# Extract model results
modd <- augment(m1) %>% mutate(index = 1:n()) 

modd$phat <- predict(m1, type="response")


par(mfrow = c(1,1)) # row-column
## plot cook's distance vs estimated probabilities
plot(modd$index,modd$.cooksd, main="cook's distance vs observation number")

# plot residuals versus fitted values
# add a lowess fit

# plot logit of estimated probabilities against predictors

```

# Penalized logistic regression

```{r,echo=T,cache=T}
load("sim_geno.rda")
load("sim_trait.rda")
# take a smaller par
sim_geno_a = sim_geno[1:500,1:100]
sim_trait_a = sim_trait[1:100]
sim_trait_a1 = as.numeric(abs(sim_trait_a)>0.5)

set.seed(123)

library(glmnet)

lga =cv.glmnet(t(sim_geno_a),sim_trait_a1,family="binomial",alpha = 1,type.measure="class")

Ma = glmnet(t(sim_geno_a),sim_trait_a1, alpha = 1, family = "binomial",lambda = lga$lambda.min)
# Display non-zero regression coefficients
coef(Ma)[coef(Ma) !=0]

## obtain p-values
library(hdi)
prj_est = lasso.proj(t(sim_geno_a), sim_trait_a1, family = "binomial")

prj_est$pval[1:10]

### Ridge regression

lgb =cv.glmnet(t(sim_geno_a),sim_trait_a1,family="binomial",alpha = 0,type.measure="class")

Mb = glmnet(t(sim_geno_a),sim_trait_a1, alpha = 0, family = "binomial",lambda = lga$lambda.min)


## obtain p-values
prj_estb = ridge.proj(t(sim_geno_a), sim_trait_a1, family = "binomial")

prj_estb$pval[1:10]

```


# Multinomial logistic regression

Please go to webpage: https://stats.idre.ucla.edu/r/dae/multinomial-logistic-regression/


The data set contains variables on 200 students. The outcome variable is `prog`, program type. The predictor variables are social economic status, `ses`, a three-level categorical variable and writing score, `write`, a continuous variable. 

```{r,echo=T,cache=T}
require(foreign)
require(nnet)
require(ggplot2)
require(reshape2)

ml <- read.dta("hsbdemo.dta")

with(ml, table(ses, prog))

# set baseline level for "prog"
ml$prog2 <- relevel(ml$prog, ref = "academic")
test <- multinom(prog2 ~ ses + write, data = ml)

# "ses = low" is the baseline and absorbed into the intercept
summary(test)

# calculate z-test value
z <- summary(test)$coefficients/summary(test)$standard.errors
z

# calculate two-sided p-values using z-tests
p <- (1 - pnorm(abs(z), 0, 1)) * 2
p

# use fitted model to predict probabilities
head(pp <- fitted(test))

```

# Poisson regression

## Settings

Definition: Poisson random variable $W$ with mean $\mu >0$:
$$\Pr(W=k) = \frac{e^{-\mu}  {\mu}^k }{k!}, \quad k=0,1,\ldots$$
Consider pairs of observations $(x_i,y_i),i=1,2,\ldots,n$, where each $y_i$ is a Poisson random variable with mean $\mu_i$ and each $x_i$ is the $i$th observation from the predictor $X$, we propose the model:
$$ \ln \mu_i = \beta_0 + \beta_1 x_i$$
for a deterministic intercept $\beta_0$ and slope $\beta_0$, where $\ln$ is the natural log. This model is fit by the procedure of *maximum likelihood estimation (MLE)*, which provides estimates $\hat{\beta}_0$ and $\hat{\beta}_1$ to $\beta_0$ and $\beta_1$, respectively.

For the case of $p$ predictors $X_1,\ldots,X_p$, we have the model:

$$ \ln \mu_i  = \beta_0 + \beta_1 x_{i1} + \ldots + \beta_p x_{ip},$$
where $x_{ij}$ is the $i$th observation of $X_j$, and $\beta_1,\ldots,\beta_p$ are the regression coefficients. MLE also is used to fit this model and provides estimates $\hat{\beta}_j,j=0,1,\ldots,p$.


When the $n$ $y_i$'s are independent from each other, their joint likelihood is 
$$\Pr(y_1=k_1,\ldots,y_n=k_n) = \frac{e^{-\mu_1}  {\mu_1}^ {k_1} }{k_1!} \times \ldots \times \frac{e^{-\mu_n}  {\mu_n}^ {k_n} }{k_n!}. $$
If $$ \ln \mu_i  = \beta_0 + \beta_1 x_{i1} + \ldots + \beta_p x_{ip},$$ then the above expression $\Pr(y_1=k_1,\ldots,y_n=k_n)$ is a function of the vector $\boldsymbol{\beta} = (\beta_0,\beta_1,\ldots,\beta_p)$ and can be denoted by $l(\boldsymbol{\beta})$. Maximization of $l(\boldsymbol{\beta})$ with respect to $\boldsymbol{\beta}$ gives an estimate $\boldsymbol{\hat{\beta}}$ of $\boldsymbol{\beta}$. 

## Example

The following contents are adopted from: https://stats.idre.ucla.edu/r/dae/poisson-regression/. 
In this example, `num_awards` is the outcome variable and indicates the number of awards earned by students at a high school in a year, `math` is a continuous predictor variable and represents students' scores on their math final exam, and `prog` is a categorical predictor variable with three levels indicating the type of program in which the students were enrolled. It is coded as 1 = "General", 2 = "Academic" and 3 = "Vocational".

```{r,echo=T,cache=T}

require(ggplot2)
require(sandwich)
require(msm)


p <- read.csv("poisson_sim.csv")
p <- within(p, {
  prog <- factor(prog, levels=1:3, labels=c("General", "Academic", 
                                                     "Vocational"))
  id <- factor(id)
})
summary(p)


ggplot(p, aes(num_awards, fill = prog)) +
  geom_histogram(binwidth=.5, position="dodge")

summary(m1 <- glm(num_awards ~ prog + math, family="poisson", data=p))


## calculate and store predicted values
p$phat <- predict(m1, type="response")

## order by program and then by math
p <- p[with(p, order(prog, math)), ]

## create the plot
ggplot(p, aes(x = math, y = phat, colour = prog)) +
  geom_point(aes(y = num_awards), alpha=.5, position=position_jitter(h=.2)) +
  geom_line(size = 1) +
  labs(x = "Math Score", y = "Expected number of awards")

```